---
title: "Mixed Models in R"
subtitle: "A Practical Introduction"
author: "Henrik Singmann (University of Zurich)<br/>Twitter: <a href='https://twitter.com/HenrikSingmann'>@HenrikSingmann</a>"
date: "June 2017"
output:
  xaringan::moon_reader:
    css: ["default", "my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
---



```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
# see: https://github.com/yihui/xaringan
# install.packages("xaringan")
# see: 
# https://github.com/yihui/xaringan/wiki
# https://github.com/gnab/remark/wiki/Markdown
options(width=110)
options(digits = 4)
```

# Mixed Models

- Mixed models are a modern class of statistical models that extend regular regression models by including random effects parameters to account for dependencies among related data points.

--

.pull-left[

### Fixed Effects
- Overall  or *population-level average* effect of specific model term (i.e., main effect, interaction, parameter) on dependent variable
- Independent of stochastic variability controlled for by random effects
- Hyptothesis tests on fixed effect interpreted as hypothesis tests for terms in standard ANOVA or regression model
- Possible to test specific hypotheses among factor levels (e.g., planned contrasts)

]

.pull-right[
### Random Effects

- *Random effect grouping factors*: Categorical variables that capture random or stochastic variability (e.g., participants, items, groups, or other hierarchical-structures).
- In experimental settings, random effects grouping factors often part of design one wants to generalize over.
- Random effects factor out idiosyncrasies of sample, thereby providing a more general estimate of the fixed effects of interest.
]

---

# Example Data

description follows...

---



---
### References Statistical Modeling:
- Singmann, H., & Kellen, D. (in press). An Introduction to Mixed Models for Experimental Psychology. In D. H. Spieler & E. Schumacher (Eds.), New Methods in Neuroscience and Cognitive Psychology. Psychology Press. http://singmann.org/download/publications/singmann_kellen-introduction-mixed-models.pdf

### References Example Data:
- Skovgaard-Olsen, N., Singmann, H., & Klauer, K. C. (2016). The relevance effect and conditionals. *Cognition*, 150, 26-36. https://doi.org/10.1016/j.cognition.2015.12.017



```{r, eval=FALSE, include=FALSE}
require(afex)
load("../exercises/ssk16_dat_preapred.rda")
```

```{r, eval=FALSE, include=FALSE}

m_full <- mixed(dv ~ c_given_a*rel_cond*dv_question +
                       (rel_cond*c_given_a|lfdn) +
                       (rel_cond*c_given_a*dv_question|le_nr),
                     dat,
                     control = lmerControl(optCtrl = list(maxfun=1e8)),
                     method = "S")

save(m_full, file = "fitted_lmms.rda", compress = "xz")
```
